---
title: Simulating Pi Using Buffon's Needle
author: ~
date: '2019-09-20'
slug: buffon-needle
categories: []
tags: []
---
```{r, echo = F, message = F, warning = F}
library(dplyr)
library(ggplot2)
```

It figures that Buffon's needle is one letter away from Buffoon's needle. You've got to be a buffoon to try to estimate $\pi$ by tossing needles at a piece of paper. Or you're me, and you're an even bigger buffoon because you used $\pi$ to simulate tossing needles at a piece of paper to estimate $\pi$. 

I first learned about Buffon's needle as a homework problem in one of my stats classes. The setup is as follows: 

Throw a needle of length $l$ at a piece of paper marked with vertical (or horizontal, it's symmetric) lines separated by width $h$. What's the probability that the needle crosses one of the lines? 

I won't go through the math here, but in the special case that $l = h$, it turns out to be $\frac{2}{\pi} \approx .637$

To do the simulation, I picked a random point for one end of the needle to land, and then a random angle for the direction of the rest of the needle. I realize this is cheating because I choose an angle between 0 and $2 \pi$, but the idea was to simulate the physical situation of tossing a needle. Nature also lets you use pi as a constant. Oh well. 

To visualize this, let's first look at just a couple tosses. All the implementation code is on my GitHub as usual. 

```{r, echo = F, message = F, warning = F}
place_needle_head <- function(dim = 3) {
  x <- runif(1, -dim, dim)
  y <- runif(1, -dim, dim)
  return(c(x,y))
}

pick_needle_angle <- function() {
  return(runif(1, 0, 2*pi))
}

get_needle_endpoints <- function(dim = 3) {
  s_point <- place_needle_head(dim)
  angle <- pick_needle_angle()
  endX = s_point[1] + cos(angle)
  endY = s_point[2] + sin(angle)
  return(data.frame(startX = s_point[1], endX = endX,
                    startY = s_point[2], endY = endY))
}

needle_hits <- function(needle_location) {
  return(floor(needle_location$startX) != floor(needle_location$endX))
}


make_needle_df <- function(n_needles = 1000, dim = 3, random_seed = F) {
  if(random_seed) {
    set.seed(random_seed)  
  }
  needle_df <- vector("list", length = n_needles)
  for (i in 1:n_needles) {
    needle_df[[i]] = get_needle_endpoints(dim)
  }
  needle_df <- bind_rows(needle_df) %>% 
    mutate(dim = dim,
           hits = floor(startX) != floor(endX))
  return(needle_df)
}

plot_needle_drop <- function(needle_df, dim = 4) {
  
  pi_approx <- round(2 / mean(needle_df$hits), 3)
  
  needle_df %>% 
    rename(`Needle Hits` = hits) %>% 
    ggplot(aes(x = startX, y = startY, xend = endX, yend = endY)) +
      geom_segment(aes(color = `Needle Hits`), size = 2) +
      labs(title = paste("Simulating", nrow(needle_df), "iterations of Buffon's Needle"),
           #subtitle = paste0("* Approximation of ",
          #                   expression(pi), " is ",
          #                   round(pi_approx, 3)),
           subtitle = bquote(pi %~~% .(pi_approx)),
           x = "", y = "") +
      scale_color_manual(values = c("blue4", "darkgoldenrod2")) +
      coord_cartesian(xlim = c(-dim, dim), ylim = c(-dim, dim)) +
      scale_x_continuous(breaks = seq(-dim, dim, 1)) +
      scale_y_continuous(breaks = seq(-dim, dim, 1)) +
      theme(aspect.ratio=1,
            panel.background = element_rect(fill = alpha("white")),
            axis.text = element_blank(),
            axis.ticks = element_blank(),
            panel.grid.major.y = element_blank(),
            panel.grid.minor = element_blank(),
            panel.grid.major.x = element_line(color = "black", size = .25))
}

theme_clean <- function () { 
    theme_bw(base_size = 11) +
   theme(panel.grid.major.x=element_blank(), panel.grid.minor.x = element_blank(),
           panel.grid.minor.y = element_blank(), panel.background = element_rect(colour = "#E1E1E1"),
           panel.border = element_blank())
}
```

```{r, echo = F, message = F, warning = F}
set.seed(13)
plot_needle_drop(make_needle_df(10))
```

Seven of the ten needles crossed a vertical line, giving us an approximation of $\pi$ of $\frac{2}{.7} \approx 2.857$.

Let's try it with a few more needles: 

```{r, echo = F, message = F, warning = F}
set.seed(3)
plot_needle_drop(make_needle_df(100))
```

Binomial trials clearly aren't a very good way to quickly estimate $\pi$, but to verify that it really will converge, let's plot the estimate of pi after each of 10,000 trials:

```{r, echo = F, message = F, warning = F}
ten_thousand_needles <- make_needle_df(10000, 3, 1313)

ten_thousand_needles %>% 
  mutate(trial = row_number(),
         hit_rate = cummean(hits),
         pi_est = 2 / hit_rate) %>% 
  filter(trial > 50) %>% 
  ggplot(aes(x = trial, y = pi_est)) +
    geom_hline(yintercept = pi, color = "darkgoldenrod2", size = 1.5) +
    geom_line(color = "blue4", size = 1.5) +
    labs(title = bquote("Buffon's Needle Converging to" ~ pi),
         x = "Trial Number",
         y = bquote("Estimate of" ~ pi)) +
    coord_cartesian(ylim = c(2.5, 4.5)) +
    theme_clean()

```

It turns out that Buffon was pretty smart.

