---
title: Playing Around with The Office Script
author: ~
date: '2020-04-04'
slug: playing-around-with-the-office-script
categories: []
tags: []
---
```{r, include=F, echo=F}
knitr::opts_chunk$set(warning=FALSE, message = FALSE)
```

This post will be a little bit different from most, in that I'm going to show all of my code.  

```{r}
file_path <- "../../static/data/office" 

theme_clean <- function() {
  theme_bw(base_size = 11) +
    theme(text = element_text(family = "Roboto"),
          panel.grid.major.x = element_blank(),
          panel.grid.minor.x = element_blank(),
          panel.grid.minor.y = element_blank(),
          panel.background = element_rect(colour = "#E1E1E1"),
          panel.border = element_blank())
}

library(scales)
library(lubridate)
library(readr)
library(glue)
library(dplyr)
library(ggplot2)
library(stringr)
```

```{r}
# Read in the data
script <- read_csv(glue("{file_path}/office_script.csv"))

# Remove things in brackets, they're not spoken words
script <- script %>%
  mutate(line_text = gsub("\\[.*?\\]", "", line_text)) %>%
  # Remove the deleted scenes
  filter(!deleted) %>%
  select(-deleted)
```

```{r}
# Fix some inconsistencies in the names
script <- script %>%
  mutate(speaker = case_when(speaker == "David" ~ "David Wallace",
                             speaker == "Deangelo" ~ "DeAngelo",
                             speaker == "Ryan Howard" ~ "Ryan",
                             speaker == "Robert" ~ "Robert California",
                             speaker == "Darly" ~ "Darryl",
                             speaker == "Bob" ~ "Bob Vance",
                             speaker %in% c("Michel", "Micahel", "Michae",
                                            "Michael [on phone]", "M ichael",
                                            "Micael", "MIchael", "Michal", "Mihael", "Miichael" ) ~ "Michael",
                             T ~ speaker))

remove_text_punc <- function(string) {
  string %>% str_replace_all(pattern = "[\n-]+|\\s+",
                      replacement = " ") %>%
      str_replace_all(pattern = "[\\.,!/?():]",
                      replacement = "") %>%
      tolower()
}



# Add on speaker group, for merging later
script <- script %>%
  mutate(unique_speaker_order = cumsum(c(TRUE, speaker[-1] != speaker[-length(speaker)])))


# Make a somewhat degenerate data frame, but that has the 
# lines someone says without someone else interrupring 
# All considered to be the same line

script_grouped_lines <- aggregate(line_text ~ speaker + unique_speaker_order, transform(script, 
    unique_speaker_order = cumsum(c(TRUE, speaker[-1] != speaker[-length(speaker)]))), 
           FUN = paste, collapse= " ")

# Merge to collapse rows of the same speaker
script <- script %>%
  select(-line_text, - speaker) %>%
  group_by(unique_speaker_order) %>%
    slice(1) %>%
  ungroup() %>%
  merge(script_grouped_lines %>%
          select(speaker, line_text, unique_speaker_order),
        by = "unique_speaker_order") %>%
  select(-unique_speaker_order)

# Remove punctuation, make things lowercase
script <- script %>%
  mutate(line_text = remove_text_punc(line_text))

# Add in the number of words per line

script <- script %>% 
  mutate(nwords = str_count(line_text, " ") + 1)

```

```{r}
script %>% 
  filter(is.na(nwords))
```

Once we've read in and cleaned the data, we're ready to look at the number of lines per character:

```{r}

speaker_line_ranking <- script %>%
  group_by(speaker) %>%
  summarize(n_lines = sum(nwords)) %>%
  arrange(-n_lines) %>%
  filter(n_lines > 1000) %>%
  ungroup() %>%
  pull(speaker)

script %>%
  group_by(speaker) %>%
  summarize(n_lines = sum(nwords)) %>%
  mutate(more_than_100 = n_lines > 1000,
         speaker_adj = ifelse(more_than_100, speaker, "Other")) %>%
  ungroup() %>%
  group_by(speaker_adj) %>%
  summarize(n_lines = sum(n_lines)) %>%
  arrange(-n_lines) %>%
  mutate(speaker = factor(speaker_adj, levels = c(speaker_line_ranking, "Other"))) %>%
  ggplot(aes(x = speaker, y = n_lines)) +
    geom_bar(stat = "identity") +
    labs(title = "Total Lines by Character",
         x = "Speaker",
         y = "Number of Lines") +
    theme_clean() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Great, let's look at how the top 12 characters' (through Kelly on the graph above) prominence changed as the seasons went on. I'm going to show this two ways, once where every character gets their own scale, so that you can easily see changes in that character, and once where every character is on the same scale, so that it's easier to see the relative proportions.  

```{r}
# Season Level proportion of lines by character
season_line_counts <- script %>%
  group_by(season) %>%
  summarize(season_lines = n()) %>%
  merge(data.frame(speaker = speaker_line_ranking[1:12]))


prevelance_by_season_plot <- script %>%
  filter(speaker %in% speaker_line_ranking[1:12]) %>%
  group_by(speaker, season) %>%
  summarize(n_lines = n()) %>%
  merge(season_line_counts, by = c("season", "speaker"), all.y = T) %>%
  mutate(n_lines = ifelse(is.na(n_lines), 0, n_lines)) %>%
  mutate(season_prop = n_lines/season_lines,
         speaker = factor(speaker, levels = speaker_line_ranking)) %>%
  ggplot(aes(x = season, y = season_prop)) +
    geom_point() +
    geom_line() +
    scale_x_continuous(breaks = 1:9) +
    labs(title = "Character Importance by Line Frequency over Time",
         x = "Season", y = "Proportion of Lines") +
    theme_clean()

prevelance_by_season_plot +
  facet_wrap(~speaker, scale= "free") 
```

```{r}
prevelance_by_season_plot +
  facet_wrap(~speaker) 
```

Michael's plot looks the same either way, but the second version makes it clear just how much he carried the early seasons of the show. Together, these plots show a kind of rise of most of the minor characters as the show moved away from, and eventually left out, Michael.  

I'll have to come back to this, there's a lot more we can do with the script. 







